\section{Introduction}

When implementing a new language, whether for research purposes or as
a realistic general purpose language, we are inevitably faced with the
problem of executing the language. Ideally, we would like execution to
be as fast as possible, and exploit known techniques from many years
of compiler research. However, it is difficult to make use of the
existing available back ends for functional langauges, such as the
STG~\cite{evalpush,stg,llvm-haskell} or ABC~\cite{abc-machine}
machines. They may be too low level, they may make assumptions about
the source language (e.g. its type system) or there may simply be no
clearly defined API. As a result, experimental languages such as
Agda~\cite{norell-thesis} have resorted to generating Haskell, using
\texttt{unsafeCoerce} to bypass the type system. Similarly,
Cayenne~\cite{cayenne-icfp} generated LML bypassing the type
checker. This is not ideal for several reasons: we cannot expect to
use the full power and optimisations of the underlying compiler, nor
can we expect it to exploit any specific features of our new source
language, such as the optimisation opportunities presented by rich
dependent type systems~\cite{brady-thesis}.

Epic is a library which aims to provide the necessary features for
implementing the back-end of a functional language --- thunks,
closures, algebraic data types, scope management, lambda lifting ---
without imposing \remph{any} design choices on the high level language
designer, other than encouraging a functional style. It
provides \remph{compiler combinators}, which guarantee that any output
code will be syntactically correct and well-scoped.  This gives a
simple method for building a compiler for a new functional language,
e.g. for experimentation with new type systems or new domain specific
languages. In this paper, we describe Epic and its API using two
example high level languages. More generally, we observe that:

\begin{enumerate}
\item Recent language and type system research has typically been
  based on extensions of existing languages, notably Haskell. While
  this makes implementation easier as it builds on an existing
  language, it discourages significant departures from the existing
  language (e.g. full dependent types). With Epic, we encourage
  more radical experiments by providing a standalone path to a realistic,
  efficient, language implementation.
\item A tool can become significantly more useful if it is embeddable in
  other systems. A language back end is no different --- by providing
  an API for Epic, we make it more widely applicable. Haskell's
  expressiveness, particularly through type classes, makes it simple
  to provide an appropriate API for describing the core language.
\item Epic's small core and clearly defined API makes it a potential
  platform for experimentation with optimisations and new back
  ends. Indeed, we avoid implementation details in
  this paper. Several implementations are possible, perhaps targetting
  .NET or the JVM as well as native code.
\end{enumerate}

\noindent
Epic was originally written as a back end for
Epigram~\cite{levitation} (the name\footnote{Coined by James McKinna}
is short for ``\textbf{Epi}gram \textbf{C}ompiler''). It is now used
by Idris~\cite{plpv11} and as an experimental back end for
Agda. It is specifically designed for reuse by other source languages.





